---
title: "Cases of Mumps In New York City"
author: "Isis Ramirez"
date: "4/16/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(astsa)
library(xts)
library(ggplot2)
library(gridExtra)
library(forecast)
library(knitr)
library(kableExtra)
library(changepoint)
require(quantmod)
require(timeSeries)
```

## Background and Key Questions

We will be looking at a time series dataset on the monthly count of mumps cases in New York City from 1928-1972.

The following questions will be addressed:   

- Is there a temporal pattern to mumps occurrences?
   
- Is the vaccine effective at preventing mumps?


## Data: Definition, Acquisition and Exploration    

The time series plot of the data shows periodic oscillations.
```{r}
## Load Data & Format ##
mumps <- read.csv("mumps_data.csv", header = TRUE)
ts.mumps <- ts(mumps[,2], start=c(1928,1), end=c(1972,6), frequency=12)

## Plot Data & Decomposition
autoplot(ts.mumps, ylab="Number of mumps Cases", 
         main="Reported Mumps Cases in New York City")
```

## Seasonality   

There are higher numbers of occurrences in late winter and early spring. 
```{r}
seasonplot(ts.mumps, year.labels=TRUE, col= 1:16, ylab="Number of Mumps Cases")
```


## Modeling Cases of Mumps   

The acf shows periodic behavior. 
```{r}
invisible(acf2(ts.mumps))
```

## Modeling Cases of Mumps   

Below are the results for a SARIMA(2,0,0)x(1,0,1)[12] with exogenous variables using trigonometric functions.
$$\small x_{t}=A[I_{pre}cos(\frac{2\pi t}{12}+\phi_{1})]+B[I_{post}cos(\frac{2\pi t}{12}+\phi_{2})]$$ 
```{r, fig.height= 3}
n <- length(ts.mumps)
t <- 1:length(ts.mumps)
w <- 1/12

PreVac <- as.numeric(time(ts.mumps)<1967)
PostVac <- as.numeric(time(ts.mumps)>1967)

cos.D <- cos(2*pi*w*t) 
sin.D <- sin(2*pi*w*t) 
Dmatrix <- cbind(cos.D*PreVac, sin.D*PreVac, cos.D*PostVac, sin.D*PostVac)

fit <- invisible(astsa::sarima(ts.mumps, p=2,d=0,q=0, P=1, Q=1, S=12, 
                        xreg=Dmatrix, no.constant=FALSE,details=F))
ttable <- round(fit$ttable[,-3],2)
rowNames <- c("ar1","ar2","sar1","sma1","intercept","PreV1","Prev2","PostV3","PostV4") 
rownames(ttable) <- rowNames
param <- kable(ttable, type='latex')

B1.PreV <- fit$ttable[6,1]
B2.PreV <- fit$ttable[7,1]
B1.PostV <- fit$ttable[8,1]
B2.PostV <- fit$ttable[9,1]

amp1 <- sqrt(B1.PreV^2+B2.PreV^2)
amp2 <- sqrt(B1.PostV^2+B2.PostV^2)

amplitudes <- as.data.frame(c(round(amp1,2), round(amp2,2)))
rowNames <- c("Pre-Vaccine (A)", "Post-Vaccine (B)")
rownames(amplitudes) <- rowNames
colNames <- c("Amplitudes")
colnames(amplitudes) <- colNames
amps <- kable(amplitudes, type='latex')

grid.arrange(tableGrob(ttable), tableGrob(amplitudes), ncol = 2)

```


## Change in Mean   

The changepoint package determined a change in mean.   
The estimated mean of the process is 437.   
```{r}
mvalue <-  cpt.mean(ts.mumps)
changePoint <- cpts(mvalue)
changeDate <- as.yearmon(time(ts.mumps))[changePoint]

m1 <- round(mvalue@param.est$mean[1],0)
m2 <- round(mvalue@param.est$mean[2],0)

plot(mvalue, ylab="Number of Mumps Cases")
abline(v=time(ts.mumps)[changePoint], lty=2)
text(time(ts.mumps)[changePoint]+1, 750, labels = changeDate,  srt=45, col="blue")
text(1928, m1+60, labels = m1, col="blue")
text(1973, m2+60, labels = m2, col="blue")
```


## Summary Conclusions

- The time series presents periodic behavior.

- Higher occurrences in late winter and spring.

- Decreased amplitude and mean after the introduction of the vaccine.
